A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-29 paper.tex

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to regard the annotations as describing
the complete set of genomic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as a gARG,
with contrasting
edge annotation schemes. (A) Immediate edge annotation, reflecting ``local''
transmission of all genetic material between parent and child. All edges span
the full genome from position 0 to 7, apart from those linking a recombinant
and its two parents (bold). (B) Sample-resolved edge annotation: the
Hudson-like simplification algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the grand MRCA \textsf{w}. Line widths reflect
the genomic span of each edge, which helps visualise the relative importance of
different transmission pathways.
}
\end{figure}

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}B) is generally much more
useful than storing the local parent-child intervals of inheritance
(Fig~\ref{fig-ancestry-resolution}A).
Indeed, storing local parent-child
inheritance intervals for edges is really only of interest because of
its direct correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG. The eARG encoding
is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to simulate or infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlikey to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o} and \textsf{v}).

@@

Given the local inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}A),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}B) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the ancestral material for a node is the
union of the ancestral intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their local inheritance and
just-computed ancestral intervals for the node.
For example, node \textsf{d} in Fig~\ref{fig-ancestry-resolution}
carries ancestral material on the interval $(0, 4]$, since it
is the left-hand parent of the recombinant \textsf{b}. The edge
joining \textsf{d} to its ancestor \textsf{h} then carries ancestral
material on $(0, 4]$ since although \textsf{d} inherits the entire interval
$(0, 7]$ from \textsf{h}, only $(0, 4]$ is ancestral to the sample.
This process is directly analogous to Hudson's simulation algorithm
(Appendix XXX).

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-in-time vs backwards-in-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
is the same key algorithm underlying recent advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, recording
the full genetic ancestry of the current population in this way is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.

>>>>

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs.
Fig~\ref{fig-ancestry-resolution} is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).

There are two distinct annotation schemes possible, and the
distinction is subtle and important.
The first scheme is to use the annotations to describe
the complete set of genomic intervals inherited by the descendant node from the
ancestor node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to describe
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as an eARG and
two gARGs with alternative edge annotation schemes.
(A) A conventional Griffiths eARG. Square nodes represent events, with
recombinations (in red) annotated with the position of a single breakpoint
along the 7-unit long chromosome.
(B) A gARG with direct-inheritance edge annotations. Oval nodes represent
genomes and annotations reflect ``local'' transmission of all genetic material
between parent and child. Edges that do not span the full genome are visually
highlighted, showing how breakpoint positions can be encoded.
(C) A gARG with sample-resolved edge annotations. The Hudson-like simplification
algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the MRCA nodes \textsf{k} (regions 0 to 2) and \textsf{p}
(regions 2 to 7). Line widths reflect the genomic span of each edge, highlighting
the relative importance of different transmission pathways.
}
\end{figure}

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
useful than storing the local parent-child intervals of inheritance
(Fig~\ref{fig-ancestry-resolution}B).
Indeed, storing local parent-child
inheritance intervals for edges is really only of interest because of
its direct correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
The eARG encoding is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

Note that this example is a sample from the so-called \emph{``big'' ARG process}
(see Appendix XXX) and has several features that we are unlikely to simulate or
infer in practice. For example, by definition the samples contain no genetic information about
nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{q} as all regions of the genome have already reached a
common ancestor separately (in nodes \textsf{k} and \textsf{p}).

@@

Given the local inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the ancestral material for a node is the
union of the ancestral intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their local inheritance intervals
and the just-computed ancestral intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries ancestral material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries ancestral
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. This process is directly analogous to Hudson's
simulation algorithm
(Appendix XXX).

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-in-time vs backwards-in-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}B) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}C)
is the same key algorithm underlying recent advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, using this approach to
record the full genetic ancestry of the current population is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.
